---
title: "How does the signature size influence the BioQC score? "
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BioQC)
library(data.table)
library(ggplot2)
```

```{r}
addNoise = function(matrix, fractionAffected=.1, stdv=2) {
   noise = matrix(rnorm(nrow(matrix)*ncol(matrix), mean=9, sd=stdv), nrow=nrow(matrix), byrow=FALSE)
   addNoise = matrix(runif(nrow(matrix)*ncol(matrix)), nrow=nrow(matrix), byrow=FALSE) < fractionAffected
   matrix = matrix + addNoise*noise
   return(matrix)
}
```

```{r}
N_SAMP = 50
N_GENES = 20000
N_SIG = 100
```

We generate a random gene expression matrix with `r N_SAMP` samples and `r N_GENES` Genes. 
We generate `r N_SIG` different random signatures of length 10 and 1000 respectively. 
The following figure shows the p-value distribution. 
```{r}


sample_mat = matrix(rlnorm(N_GENES * N_SAMP, mean=9, sd=2), N_GENES, N_SAMP)
sample_mat = addNoise(sample_mat, stdv=15)
sig_list10 = lapply(1:N_SIG, function(unused) {return(sample(1:(N_GENES), 10))})
sig_list1000 = lapply(1:N_SIG, function(unused) {return(sample(1:(N_GENES), 1000))})

bioqc_res_10 = data.table(melt(absLog10p(wmwTest(sample_mat, sig_list10, valType="p.greater"))))
bioqc_res_1000 = data.table(melt(absLog10p(wmwTest(sample_mat, sig_list1000, valType="p.greater"))))
invisible(bioqc_res_10[,sl:="10"])
invisible(bioqc_res_1000[,sl:="1000"])
res_molten = rbind(bioqc_res_10, bioqc_res_1000, fill=TRUE)
ggplot(res_molten, aes(x=value)) + geom_density(aes(color=sl))
```

What if the signatures occur towards the top of the list? 
As expected, large signatures are more potent than small ones. 
```{r}
sample = 1:20000
signature1 = c(1)
signature10 = c(1:10)
signature100 = c(1:100)

p1 = -log10(wmwTest(sample, signature1, valType = "p.greater"))
p10 = -log10(wmwTest(sample, signature10, valType = "p.greater"))
p100 = -log10(wmwTest(sample, signature100, valType = "p.greater"))
```

## Signature positioning
```{r}
sample = 1:20000
signature_0 = 1:101
signature_25 = 4050:5050
signature_50 = 9950:10050
signature_75 = 14950:15050
signature_75_10 = seq(14500, 15500, 10) # signature spread has no effect -> median could suffice as correction value
signature_100 = 19900:20000

wmwTest(sample, signature_0, valType = "p.greater")
wmwTest(sample, signature_25, valType = "p.greater")
wmwTest(sample, signature_50, valType = "p.greater")
wmwTest(sample, signature_75, valType = "p.greater")
wmwTest(sample, signature_75_10, valType = "p.greater")
wmwTest(sample, signature_100, valType = "p.greater")

```

## How does it look like for other statistical tests? 
### t-test
```{r}
do.test <- function(senseMat, tissueInd, test=t.test, alternative="greater") {
   bgInds = setdiff(1:nrow(senseMat), tissueInd)
   res = apply(senseMat, 2, function(col) {
     gs = col[tissueInd]
     bg = col[bgInds]
     return(test(gs, bg, alternative=alternative)$p.value)
   })
   return(res)
}

t_res10 = data.table(melt(lapply(sig_list10, function(signature) {
  absLog10p(do.test(sample_mat, signature, test=t.test))
})))
t_res1000 = data.table(melt(lapply(sig_list1000, function(signature) {
  absLog10p(do.test(sample_mat, signature, test=t.test))
})))
invisible(t_res10[,sl:="10"])
invisible(t_res1000[,sl:="1000"])
res_molten = rbind(t_res10, t_res1000)
ggplot(res_molten, aes(x=value)) + geom_density(aes(color=sl))

```


### KS-test
```{r}

ks_res10 = data.table(melt(lapply(sig_list10, function(signature) {
  absLog10p(do.test(sample_mat, signature, test=ks.test))
})))
ks_res1000 = data.table(melt(lapply(sig_list1000, function(signature) {
  absLog10p(do.test(sample_mat, signature, test=ks.test))
})))
invisible(ks_res10[,sl:="10"])
invisible(ks_res1000[,sl:="1000"])
res_molten = rbind(ks_res10, ks_res1000)
ggplot(res_molten, aes(x=value)) + geom_density(aes(color=sl))

```