---
title: "R Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BioQC)
library(data.table)
library(ggplot2)
library(assertthat)
library(RColorBrewer)
library(scales)
```

```{r}
addNoise = function(matrix, fractionAffected=.1, stdv=2) {
   noise = matrix(rnorm(nrow(matrix)*ncol(matrix), mean=20, sd=stdv), nrow=nrow(matrix), byrow=FALSE)
   addNoise = matrix(runif(nrow(matrix)*ncol(matrix)), nrow=nrow(matrix), byrow=FALSE) < fractionAffected
   matrix = matrix + addNoise*noise
   return(matrix)
}
```


```{r}
SIG_SIZE = 500
OVERLAP = 250
N_SIG = 100
# N_SAMP = 100
N_GENES = 20000
sig_list_liver = lapply(1:N_SIG, function(unused) {return(sample(1:(N_GENES), SIG_SIZE))})
sig_list_kidney = lapply(1:N_SIG, function(i) {return(sample(sig_list_liver[[i]], OVERLAP))})
sig_list_kidney = lapply(1:N_SIG, function(i) {
  return(c(sample((1:N_GENES)[-sig_list_liver[[i]]], SIG_SIZE-OVERLAP), sig_list_kidney[[i]]))
})
invisible(lapply(1:N_SIG, function(i) {
  assert_that(length(intersect(sig_list_kidney[[i]], sig_list_liver[[i]])) == OVERLAP)
}));

# make random samples
sample_mat = matrix(rlnorm(N_GENES * N_SAMP, mean=9, sd=2), N_GENES, N_SAMP)
sample_mat_overexpressed = matrix(rlnorm(N_GENES * N_SAMP, mean=12, sd=2), N_GENES, N_SAMP)

# make liver 'genes' overexpressed
for(i in 1:N_SIG) {
  sample_mat[sig_list_liver[[i]],i] = sample_mat_overexpressed[sig_list_liver[[i]],i]
}

bioqc_res_liver = data.table(melt(diag(absLog10p(wmwTest(sample_mat, sig_list_liver, valType="p.greater")))))
bioqc_res_kidney = data.table(melt(diag(absLog10p(wmwTest(sample_mat, sig_list_kidney, valType="p.greater")))))
bioqc_res_liver[,sl:="liver"]
bioqc_res_kidney[,sl:="kidney"]
res_molten = rbind(bioqc_res_liver, bioqc_res_kidney, fill=TRUE)
ggplot(res_molten, aes(x=value)) + geom_density(aes(color=sl))
```





The 'expected pvalue' only depends on the overlap and the size of the overlapping signature (in this case, on the kidney signature and the overlap with the liver signatures. )

In other words, the relation of the overlap to the signature. 

We simulate two scencarios: 

* worst-case: the overlap are the top ranking genes in the sample
* average-case: not all genes in the liver signature are over-expressed. The overlap is a random subsample of the genes in the liver signature. 

## worst-case

```{r}
SIG_SIZE = 50
N_GENES = 20000

bioqc_res = lapply(1:SIG_SIZE, function(overlap) {
  sig_liver = 1:SIG_SIZE
  sample = c(500:(500+overlap), rnorm(N_GENES-overlap, mean = 9, sd = 2))
  return(absLog10p(wmwTest(sample, sig_liver)))
})

ggplot(melt(bioqc_res), aes(x=L1, y=value)) + geom_line()
```

    
```{r}
N_GENES = 40000
overlap = seq(.01, .2, .01)
names(overlap) = overlap
sig_length  = seq(20, 1000, 20)
names(sig_length) = sig_length


bioqc_res = lapply(overlap, function(ol) {
  lapply(sig_length, function(sl) {
    sig_liver = 1:(ol*sl)
    sample = c(500+sig_liver, rnorm(N_GENES-length(sig_liver), mean=9, sd=2))
    return(absLog10p(wmwTest(sample, sig_liver)))
  })
})

bioqc_res_melt = data.table(melt(bioqc_res))
colnames(bioqc_res_melt) = c("pvalue", "sig_length", "percent_overlap")
bioqc_res_melt[,sig_length:=factor(sig_length, levels=unique(as.character(sig_length)))]

hm.palette <- colorRampPalette(rev(brewer.pal(11, 'Spectral')), space='Lab')  
ggplot(bioqc_res_melt, aes(x=sig_length, y=percent_overlap)) + geom_tile(aes(fill=pvalue)) +   
  scale_fill_gradientn(colours = hm.palette(100), limits=c(0, 30), oob=squish) 

```


## average case
```{r}
N_GENES = 20000
FRACTION_OVEREXPRESSED = .5
overlap = seq(.01, .2, .01)
names(overlap) = overlap
sig_length  = seq(20, 1000, 20)
names(sig_length) = sig_length


bioqc_res = lapply(overlap, function(ol) {
  lapply(sig_length, function(sl) {
    sig_liver = 1:(ol*sl)
    sample = rnorm(N_GENES, mean=9, sd=2)
    sample_oe = rnorm(N_GENES, mean=12, sd=2)
    oe_inds = sample(sig_liver, floor(FRACTION_OVEREXPRESSED*length(sig_liver)))
    sample[oe_inds] = sample_oe[oe_inds] 
    return(absLog10p(wmwTest(sample, sig_liver)))
  })
})

bioqc_res_melt = data.table(melt(bioqc_res))
colnames(bioqc_res_melt) = c("pvalue", "sig_length", "percent_overlap")
bioqc_res_melt[,sig_length:=factor(sig_length, levels=unique(as.character(sig_length)))]

hm.palette <- colorRampPalette(rev(brewer.pal(11, 'Spectral')), space='Lab')  
ggplot(bioqc_res_melt, aes(x=sig_length, y=percent_overlap)) + geom_tile(aes(fill=pvalue)) +   
  scale_fill_gradientn(colours = hm.palette(100), limits=c(0, 30), oob=squish) 

```